Clearance Project - 4 - PhysChem Clustering

### 2025-03-20 data-raw/ClearanceProject-4-PhysChemClustering.Rmd
Ensured in-progress analysis scripts were able to run sequentially. Further refinement expected in the future.

```{r data_loading}
target_chem_properties <- readRDS("data-raw/ctxR_3comp2_cheminfo.rds")
new_pk <- readRDS("data-raw/ratCVT_wHTTK_reformulated.rds")
red_assay_res <- readRDS("data-raw/2019-05-06-REDassay_HTTK.rds")
```

@@

target_chem_pw_full %>%
  ggplot(aes(x = forcats::fct_reorder(dtxsid, logkow))) +
  geom_pointrange(aes(y = logkow, ymin = logkow - logkow_sd,
                      ymax = logkow + logkow_sd), size = 0.1) +
  facet_grid(cols = vars(type_logkow)) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
target_chem_pw_full %>%
  ggplot(aes(x = logkow)) +
  geom_histogram(color = "grey50", fill = "grey5",
                 binwidth = 1) +
  theme_bw()

@@

# Make a matrix of all properties so far
tcs_mat <- target_chem_summary %>% 
  select(dtxsid, monoisotopicMass, Affinity.Kd.Med,
         H_donors:Rot_Bonds, logkow, apKa, bpKa, logD_7.4,
         starts_with("CYP")) %>%
  distinct()
tcs_mat <- column_to_rownames(tcs_mat, var = "dtxsid") %>% as.matrix()

# z-score all properties
# Margin = 2 is columns
tcs_mat_z <- apply(tcs_mat, MARGIN = 2, FUN = scale)
rownames(tcs_mat_z) <- rownames(tcs_mat)
summary(tcs_mat_z[, 1])
summary(tcs_mat_z[(rownames(tcs_mat_z) %in% target_chems), 1])
tcs_mat_z[is.na(tcs_mat_z)] <- 0
pca_tcs <- prcomp(tcs_mat_z, scale = FALSE)

@@

ggplot(data = pca_df %>% mutate(Target = (Chemical %in% target_chems)) %>%
         left_join(target_chem_summary, by = join_by(Chemical == dtxsid)),
       aes(x = PC1, y = PC2)) +
  geom_point(aes(color = TPSA), alpha = 0.8) +
  scale_color_viridis_c(option = "magma") +
  facet_wrap(~Target) +
  theme_bw()

@@

chem_rell_label <- total_rel_evaluation_df %>% distinct(Chemical, ll_label) %>%
  filter(ll_label %in% c("Restrictive", "Nonrestrictive"))

# Molecular descriptors in bins
bin_size <- 10
logP_plot <- target_chem_summary %>%
  inner_join(chem_rell_label, by = join_by(dtxsid == Chemical)) %>%
  distinct(dtxsid, logkow) %>% 
  left_join(fup_df, by = join_by(dtxsid == Chemical)) %>%
  mutate(logP_bin = factor(logkow %/% bin_size * bin_size)) %>%
  ggplot(aes(x = logP_bin, y = Funbound.plasma)) +
  labs(x = "logKow", y = "fup") +
  ggbeeswarm::geom_beeswarm() +
  theme_bw()

target_chem_summary %>%
  inner_join(chem_rell_label, by = join_by(dtxsid == Chemical)) %>%
  distinct(dtxsid, logkow, TPSA, Rot_Bonds, ll_label) %>% 
  pivot_longer(cols = c(logkow, TPSA),
               names_to = "Parameter", values_to = "Value") %>%
  left_join(fup_df, by = join_by(dtxsid == Chemical)) %>%
  ggplot(aes(x = Value, y = Funbound.plasma)) +
  labs(x = "", y = "fup") +
  facet_grid(cols = vars(Parameter),
             rows = vars(ll_label),
             scales = "free_x") +
  geom_point() +
  theme_bw()

>>>>

```{r data_loading}
target_chem_properties <- readRDS("data-raw/RCdata/ctxR_3comp2_cheminfo.rds")
new_pk <- readRDS("data-raw/ratCVT_wHTTK_reformulated.rds")
red_assay_res <- readRDS("data-raw/2019-05-06-REDassay_HTTK.rds")
load("data-raw/CYP_Sipes2017.RData")
```

@@

target_chem_properties %>%
  ggplot(aes(x = forcats::fct_reorder(dtxsid, logkow))) +
  geom_pointrange(aes(y = logkow, ymin = logkow - logkow_sd,
                      ymax = logkow + logkow_sd), size = 0.1) +
  # facet_grid(cols = vars(type_logkow)) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
target_chem_pw_full %>%
  ggplot(aes(x = logkow)) +
  geom_histogram(color = "grey50", fill = "grey5",
                 binwidth = 1) +
  theme_bw()

@@

# Make a matrix of all properties so far
tcs_mat <- target_chem_summary %>% 
  select(dtxsid, monoisotopicMass, logkow:Affinity.Kd.Med,
         starts_with("CYP")) %>%
  distinct()
tcs_mat <- column_to_rownames(tcs_mat, var = "dtxsid") %>% as.matrix()

# z-score all properties
# Margin = 2 is columns
tcs_mat_z <- apply(tcs_mat, MARGIN = 2, FUN = scale)
rownames(tcs_mat_z) <- rownames(tcs_mat)
summary(tcs_mat_z[, 1])
summary(tcs_mat_z[(rownames(tcs_mat_z) %in% target_chems_human), 1])
tcs_mat_z[is.na(tcs_mat_z)] <- 0
pca_tcs <- prcomp(tcs_mat_z, scale = FALSE)

@@

ggplot(data = pca_df %>% mutate(Target = (Chemical %in% target_chems_human)) %>%
         left_join(target_chem_summary, by = join_by(Chemical == dtxsid)),
       aes(x = PC1, y = PC2)) +
  geom_point(aes(color = TPSA), alpha = 0.8) +
  scale_color_viridis_c(option = "magma") +
  facet_wrap(~Target) +
  theme_bw()

@@

chem_rell_label <- this_rel_evaluated_df %>% distinct(Chemical, ll_label) %>%
  filter(ll_label %in% c("Restrictive", "Nonrestrictive"))

fup_df <- rph_param_df %>% distinct(Chemical, Funbound.plasma)
# Molecular descriptors in bins
bin_size <- 10
logP_plot <- target_chem_summary %>%
  inner_join(chem_rell_label, by = join_by(dtxsid == Chemical)) %>%
  distinct(dtxsid, logkow) %>% 
  left_join(fup_df, by = join_by(dtxsid == Chemical)) %>%
  mutate(logP_bin = factor(logkow %/% bin_size * bin_size)) %>%
  ggplot(aes(x = logP_bin, y = Funbound.plasma)) +
  labs(x = "logKow", y = "fup") +
  ggbeeswarm::geom_beeswarm() +
  theme_bw()
